\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-16 coq-example.tex
Save. Add Recursor Rules in metatheory.

We provide two examples here for illustration and evaluation.

\subsection{Type Safety for STLC}
We start with the very basic type safety proof of STLC, largely adapted from Software Foundation, which is also one of the primal motivating example of this project.

Our emphasis here is that, we singled out some programming language features from STLC and prove the type safety for each feature separately, following the style of the examples in MTC\citep{delaware2013,forsta2020}. And we use a \textit{not-yet-completely-defined} \textbf{mix-in} feature to mix the semantic and properties of two programming language feature -- product and boolean -- with vanilla STLC. This way, we can say precisely, \textit{a programming language feature itself is a piece of data/inheritance judgement/inherited family}. Compared to MTC, our example uses small-step operational semantic by exploiting the extensible inductive type, and most of the proof is directly adapted from the one in Software Foundation, resulting a more accessible proof. 
% In related work, Coq/Metatheory a la carte/Tion embedding can be emphasized as a more "semantic approach" because they encode the meaning using a special design pattern (for example, open recursive inductive type for extension), compared to our more syntactic approach. Their advantage is the transferability of this technique accross different proof assistants, and their disadvantage is that their approach are less accessible and unfriendly to amateur Coq users -- which can be reflected from the distance of their approach and text-book Software Foundation proof.

% We have problem on inversion lemma. Check if it is the same problem as MTC
We use `Closing Fact' to state and prove the inversion lemma instead of using the extensible proving mechanism `FTheorem'. The reason is that 1. it introduce much less boilerplate code because the proof for these inversion lemma should be just simple case analysis and we should rely on Coq to auto-generate them; 2. it shouldn't bother us in the future because any extension on the syntax should still satisfy these inversion lemmas; 3. most importantly, we believe this inversion ``lemma'' should be part of the definition of the syntax instead of considering the syntax as a mere concrete inductive type. We should postulate this inversion ``lemma'' like \textit{a constraint} and post-hoc-ly verify that our inductive definition did satisfy the constraint, which is exactly what we expect from `Closing Fact'.

@@

>>>>

We provide two examples here for illustration and evaluation of our Coq plugin.

\subsection{Type Safety for STLC}
We start with the very basic type safety proof of STLC, largely adapted from Software Foundation\cite{pierce2014software}, which is also one of the primal motivating example of this project.

Our emphasis here is that, we singled out some programming language features from STLC and prove the type safety for each feature separately, following the style of the examples in MTC\citep{delaware2013,forsta2020}. And we use a \textit{not-yet-formally-defined} \textbf{mix-in} feature to mix the semantic and properties of two programming language feature -- product and boolean -- with vanilla STLC. This way, we can say precisely, \textit{a programming language feature itself is a piece of data/inheritance judgement/inherited family}. Compared to MTC, our example uses small-step operational semantic by exploiting the extensible inductive type, and most of the proof is directly adapted from the one in Software Foundation, resulting a more accessible proof. 
% In related work, Coq/Metatheory a la carte/Tion embedding can be emphasized as a more "semantic approach" because they encode the meaning using a special design pattern (for example, open recursive inductive type for extension), compared to our more syntactic approach. Their advantage is the transferability of this technique accross different proof assistants, and their disadvantage is that their approach are less accessible and unfriendly to amateur Coq users -- which can be reflected from the distance of their approach and text-book Software Foundation proof.

% We have problem on inversion lemma. Check if it is the same problem as MTC
We use "Closing Fact" to state and prove the inversion lemma instead of using the extensible proving mechanism "FTheorem". The reasons are that (1) it introduce much less boilerplate code because the proof for these inversion lemma should be just simple case analysis and we should rely on Coq to auto-generate them; (2) it shouldn't bother us in the future because any extension on the syntax should still satisfy these inversion lemmas; (3) most importantly, we believe this inversion ``lemma'' should be part of the definition of the syntax instead of considering the syntax as a mere concrete inductive type. We should postulate this inversion ``lemma'' like \textit{a constraint} and post-hoc-ly verify that our inductive definition did satisfy the constraint, which is exactly what we expect from "Closing Fact".

Our formulation for bare STLC is around 400+ LOCs; the two families implementing product and boolean features both take around 300+ LOCs. 
% need comparison with the original implementation

The biggest difference in the proof script comes from the fact that we are handling ``extensible'' inductive type instead of real inductive type, and thus inversion tactic is not working and we have to manually lift these tactic into inversion lemma. This cost a mild code bloat.

Another difference comes from the fact that we need to use "Closing Fact" to manually verify the computational rules for each recursor and use tactic to ``run'' the recursor by repeatedly "rewrite" using those verified computational rules. This part is possible to be automatically handled by the plugin, but still it will result propositional computational rules.

@@

We construct four families, in one inheritance chain.

The first family Lang define the big step operator (using fuel) of a while language, with expression. But it postulates the computational information of State used by Lang. It takes about 200+ LOCs.

The second family LangwAbs define the abstract interpreter for it, together with a bunch of postulates, to prove the correctness of the abstract interpreter. It takes about 300+ LOCs.

The third Family LangMore extend LangwAbs with nat constant and addition, and if-then-else control flow, of course retaining all the soundness theorem. It takes about 200+ LOCs.

The fourth family LangRun instantiate some of the postulates of the LangMore and recover computational information of Lang. The concrete interpreter is doing as expected; the abstract interpreter is expected to act as a type-analyzer. It takes about 200+ LOCs.

And at the end, we compile LangRun into module and use Coq to run it on some simple queries.

### 2022-06-16 syntax.tex
Save. Add Recursor Rules in metatheory.

$$
\Rule[name=WSig Proj]
{\goodWSig{\Gamma}{s}{n} \quad j < n}
{\goodType{\Gamma}{\pi^j_1 s}{i} \quad \goodType{\Gamma, \pi^j_1 s}{\pi^j_2  s}{i}}
\quad
\Rule[name=Ind Sig]
{\goodWSig{\Delta}{\sigma}{n}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodWSig{\Gamma}{\sigma[\gamma]}{n}
  \quad \goodType{\Gamma}{W (\sigma[\gamma]) \equiv (W \sigma)[\gamma]}{i}}
$$

$$
\Rule[name=Ind Type]
{\goodTerm{\Gamma}{T}{\bW \sigma}}
{\goodTerm{\Gamma}{"W" \ T}{\cU}}
\quad
\Rule[name=Ind Term]
{\goodTerm{\Gamma}{T}{\bW \sigma}
  \quad \goodTerm{\Gamma}{a}{\pi^j_1\sigma}
  \quad \goodTerm{\Gamma, \pi^j_2\sigma[(id, a)]}{b}{El~("W"~T)}}
{\goodTerm{\Gamma}{"Wsup" \ T \ a \ b}{El~("W"~T)} }
$$

@@

Of course we need to put an eye on the substitution law for each piece of the syntax. Here $\bW \sigma$ is a large singleton type \cite{stone2000}, and it only has one "element", the inductive type itself, inside. We formulate in this cumbersome way because we want to have inductive type to be a field element and exposing the constructors at the type level for later \textit{pattern-matching coverage checking} (i.e. if an inductive type as field element has type $\cU$, then no information about the inductive type is exposed).\YZ{What is the distinction of 'W T' vs 'El (W T)'?}\EDJreply{(W T) is a term inside the type universe, El (W T) is really a type. They are basically the same thing but (W T) cannot be at the type position in a judgement, i.e. k |- q : (W T) is not allowed, we only have k |- q : El (W T). This is kind of standard notation (at least in both Sterling and Kaposi) so I will avoid explaining in the main text if you think ok?}\YZreply{I think explanation is warranted; readers unfamiliar with that line of work may find it confusing. BTW, is the elimination rule for "El (W T)" omitted?}\EDJreply{Please check the following paragraph for the explanation for El, and also Univ-1 is the elimination rule for "El T" in general. I just didn't write out the two equations c (El T) = T, El (c T) = T}. We follow the style of \textit{type universe à la Tarski} and distinguish types from their \textit{codes or names}, and thus we use $El\ T$ as the type given type name $T : \cU$ -- in these cases, $T$ is not allowed to locate at type position (i.e. at the right-most position of a term judgement) but $El \ T$ is.

In our meta-theory, we only formulate non-dependent eliminator, i.e. the
\textit{recursor}. Each recursor is constructed by a linkage containing
functions/handlers/pattern-matching cases dealing with each
constructor. Thus code reuse for recursor is delegated to linkage
inheritance---to reuse one particular pattern-matching case,
we just inherit the function/handler. This greatly simplifies the
meta\-theory development.

@@

Intuitively, both of the $\mu^+, \nu^+$ rules should be motivated by the implementation of our linkage, so an obvious question is what is this sealing judgement for?

@@

Of course, any fields using "Wsup" and "Wrec" directly relies on the $\bW(\sigma)$ and thus non-inheritable -- in these cases, "Wsup" can be understood as concrete constructors and "Wrec" are \textit{exhaustiveness checking}. These fields together with inductive types are the one sealing responsible for hiding. After compilation of linkage type into existential type, we can easily hide the \textit{concrete} definition of the inductive type while exposing all the fields using "Wsup" and "Wrec" like an abstract interface. Apparently, future fields relies on this abstract interface will have no problem on inheritance. \YZ{Does coverage checking mean exhaustiveness checking?}\EDJreply{Yes. The one make sure every case is handled in pattern-matching. Do you want me to replace the terminology using exhaustiveness checking? }\YZreply{Yes, I think exhaustiveness is standard parlance.}
Of course, since "Wsup" and "Wrec" directly rely on the concrete inductive definition, we will have to override these fields whenever inductive type is overridden. For "Wsup" we only need to override each constructor and for "Wrec" it is just doing exhaustiveness checking inside each children family. The recursor handlers however, can be safely inherited.
\YZ{I suppose the terminology 'sealing' is a reference to ML modules? In
ML, sealing has the connotation of hiding things behind an existential
type. So using the same word may cause the reader to have false
presumptions about what is going on here.}\EDJreply{You are right. Our "Sealing" is not hiding a particular field into an existential type, but our seal is make several fields together into an "existential type". Basically it hides the concrete definition of inductive type so that future fields doesn't have to be defined upon that.}\EDJreply{I will resolve this comment once I add the throughout example and you find it clear -- because I haven't changed any text}

>>>>

$$
\Rule[name=WSig-Proj]
{\goodWSig{\Gamma}{s}{n} \quad j < n}
{\goodType{\Gamma}{\pi^j_1 s}{i} \quad \goodType{\Gamma, \pi^j_1 s}{\pi^j_2  s}{i}}
\quad
\Rule[name=Ind-Sig]
{\goodWSig{\Delta}{\sigma}{n}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodWSig{\Gamma}{\sigma[\gamma]}{n}
  \quad \goodType{\Gamma}{W (\sigma[\gamma]) \equiv (W \sigma)[\gamma]}{i}}
$$

$$
\Rule[name=Ind-Type]
{\goodTerm{\Gamma}{T}{\bW \sigma}}
{\goodTerm{\Gamma}{"W" \ T}{\cU}}
\quad
\Rule[name=Ind-Term]
{\goodTerm{\Gamma}{T}{\bW \sigma}
  \quad \goodTerm{\Gamma}{a}{\pi^j_1\sigma}
  \quad \goodTerm{\Gamma, \pi^j_2\sigma[(id, a)]}{b}{El~("W"~T)}}
{\goodTerm{\Gamma}{"Wsup" \ T \ a \ b}{El~("W"~T)} }
$$

@@

Of course we need to put an eye on the substitution law for each piece of the syntax. Here $\bW \sigma$ is a large singleton type \cite{stone2000}, and it only has one ``element'', the inductive type itself, inside. We formulate in this cumbersome way because we want to have inductive type to be a field element and exposing the constructors at the type level for later \textit{pattern-matching coverage checking} (i.e. if an inductive type as field element has type $\cU$, then no information about the inductive type is exposed).\YZ{What is the distinction of 'W T' vs 'El (W T)'?}\EDJreply{(W T) is a term inside the type universe, El (W T) is really a type. They are basically the same thing but (W T) cannot be at the type position in a judgement, i.e. k |- q : (W T) is not allowed, we only have k |- q : El (W T). This is kind of standard notation (at least in both Sterling and Kaposi) so I will avoid explaining in the main text if you think ok?}\YZreply{I think explanation is warranted; readers unfamiliar with that line of work may find it confusing. BTW, is the elimination rule for "El (W T)" omitted?}\EDJreply{Please check the following paragraph for the explanation for El, and also Univ-1 is the elimination rule for "El T" in general. I just didn't write out the two equations c (El T) = T, El (c T) = T}. We follow the style of \textit{type universe à la Tarski} and distinguish types from their \textit{codes or names}, and thus we use $El\ T$ as the type given type name $T : \cU$ -- in these cases, $T$ is not allowed to locate at type position (i.e. at the right-most position of a term judgement) but $El \ T$ is.

The formulation of our recursor is closely related to the linkage, so we will postpone their description.

@@

Since our recursor handlers are closely related to linkages, we can finally talk about the rules of recursor now.

$$
\Rule[name=Hdler]
{\goodType{\Gamma}{A}{}
\quad \goodType{\Gamma, A}{B}{}
\quad \goodType{\Gamma}{R}{}
}
{\goodType{\Gamma}{"RecSig1"~A~B~R \equiv (\Pi A (\Pi (\Pi B (R[\pi_1\circ\pi_1])) R[\pi_1\circ\pi_1]) )}{}}
\quad 
\Rule[name=Hdlers]
{\goodWSig{\Gamma}{\sigma}{n}
\quad \goodType{\Gamma}{R}{}
}
{\goodSig{\Gamma}{"RecSig"~\sigma~R}{}}
$$
$$
\Rule[name=Hdler-Proj]
{ \goodWSig{\Gamma}{\sigma}{N}
\quad \goodTerm{\Gamma}{o}{\cL("RecSig"~\sigma~R)}
\quad n < N
}
{\goodTerm{\Gamma}{"prjR"~n~o}{("RecSig1"~(\pi_1^n \sigma)~(\pi_2^n \sigma)~R)[\pi_1]}}
$$
$$
\Rule[name=Rec-Constr]
{ \goodWSig{\Gamma}{\sigma}{N}
\quad \goodType{\Gamma}{T}{\bW \sigma}
\quad \goodType{\Gamma}{R}{}
\quad \goodTerm{\Gamma}{r}{\cL("RecSig"~\sigma~R)}
\quad \goodTerm{\Gamma}{w}{El~("W"~T)}
}
{\goodTerm{\Gamma}{"Wrec"~r~w}{R}}
$$

In our meta-theory, we only formulate non-dependent eliminator, i.e. the
\textit{recursor}. Each recursor is constructed by a linkage containing
functions/handlers/pattern-matching cases dealing with each
constructor. Thus code reuse for recursor is delegated to linkage
inheritance---to reuse one particular pattern-matching case,
we just inherit the function/handler. This greatly simplifies the
metatheory development.

Intuitively \ruleref{Hdler} is indicating the type of recursor handler for one specific constructor, and \ruleref{Hdlers} is the linkage type of the bundle of all recursor handlers of one given inductive type. We omit the definition of "RecSig" but \ruleref{Hdler-Proj} indicates its (apparent) internal definition. "prjR" is also the handler selection from the handler linkage, and is helpful for the formulation of $\beta$-rule of "Wrec" (omitted here). Finally \ruleref{Rec-Constr} is the used to construct a concrete recursor given a specific linkage of handlers. All these rules satisfy substitution laws and omitted here.

Now, one missing piece for linkage is the sealing judgement, and both of the $\mu^+, \nu^+$ rules relies on that -- then what is this sealing judgement for?

@@

Of course, any fields using "Wsup" and "Wrec" directly relies on the $\bW(\sigma)$ (can be spotted from \ruleref{Ind-Term} and \ruleref{Rec-Constr}) and thus non-inheritable -- in these cases, "Wsup" can be understood as concrete constructors and "Wrec" are \textit{exhaustiveness checking}. These fields together with inductive types are the one sealing responsible for hiding. After compilation of linkage type into existential type, we can easily hide the \textit{concrete} definition of the inductive type while exposing all the fields using "Wsup" and "Wrec" like an abstract interface. Apparently, future fields relies on this abstract interface will have no problem on inheritance. \YZ{Does coverage checking mean exhaustiveness checking?}\EDJreply{Yes. The one make sure every case is handled in pattern-matching. Do you want me to replace the terminology using exhaustiveness checking? }\YZreply{Yes, I think exhaustiveness is standard parlance.}
Of course, since "Wsup" and "Wrec" directly rely on the concrete inductive definition, we will have to override these fields whenever inductive type is overridden. For "Wsup" we only need to override each constructor and for "Wrec" it is just doing exhaustiveness checking inside each children family. The recursor handlers however, can be safely inherited.
\YZ{I suppose the terminology 'sealing' is a reference to ML modules? In
ML, sealing has the connotation of hiding things behind an existential
type. So using the same word may cause the reader to have false
presumptions about what is going on here.}\EDJreply{You are right. Our "Sealing" is not hiding a particular field into an existential type, but our seal is make several fields together into an "existential type". Basically it hides the concrete definition of inductive type so that future fields doesn't have to be defined upon that.}\EDJreply{I will resolve this comment once I add the throughout example and you find it clear -- because I haven't changed any text}
